# Data Structures

A data structure is a format for organizing and storing data. The structure is designed so that data can be accessed and worked with in specific ways. Statistical software and programming languages have methods (or functions) designed to operate on different kinds of data structures.

This chapter's focus is on data structures. To help initial understanding, the data in this chapter will be relatively modest in size and complexity. The ideas and methods, however, generalize to larger and more complex data sets.

The base data structures in R are vectors, matrices, arrays, data frames, and lists. The first three, vectors, matrices, and arrays, are *homogeneous*, meaning that all elements are required to be of the same type (e.g., all numeric or all character). Data frames and lists are *heterogeneous*, allowing elements to be of different types (e.g., some elements of a data frame may be numeric while other elements may be character). These base structures can also be organized by their dimensionality, as shown in Table \@ref(tab:dataStructures).

```{r, include = FALSE}
dimension <- c("1", "2", "N")
homogeneous <- c("Atomic Vector", "Matrix", "Array")
heterogeneous <- c("List", "Data Frame", "")
dat <- data.frame(dimension, homogeneous, heterogeneous)
names(dat) <- c("Dimension", "Homogeneous", "Heterogeneous")
```

```{r dataStructures, echo = FALSE}
library(knitr)
kable(
  dat, booktabs = TRUE, 
  caption = "Dimension and Type Content of Base Data Structures in R."
)
```

R has no scalar types (0-dimensional). Individual numbers or strings are actually vectors of length one.

An efficient way to understand what comprises a given object is to use the `str()` function. `str()` is short for structure and prints a compact, human-readable description of any R data structure. For example, in the code below, we prove to ourselves that what we might think of as a scalar value is actually a vector of length one. 

```{r}
a <- 1
str(a)
is.vector(a)
length(a)
```

Here we assigned `a` the scalar value one. The `str(a)` prints `num 1`, which says `a` is numeric of length one. Then just to be sure we used the function `is.vector()` to test if `a` is in fact a vector. Then, just for fun, we computed the length of `a` which again returns one. There are a set of similar logical tests for the other base data structures, e.g., `is.matrix()`, `is.array()`, `is.data.frame()`, and `is.list()`. These will all come in handy as we encounter different R objects.

## Vectors

Think of a vector^[Technically the objects described in this section are "atomic" vectors (all elements of the same type), since lists are also actually vectors. This will not be an important issue in this course, and the shorter term vector will be used for atomic vectors.] as a structure to represent one variable in a data set. For example a vector might hold the DBH, in inches, of six trees in a data set, and another vector might hold the species of those six trees. The `c()` function in R is useful for creating vectors and for modifying existing vectors. Think of `c` as standing for "combine"" or "concatenate."

```{r, tidy = FALSE}
dbh <- c(20, 18, 13, 16, 10, 14)
dbh
spp <- c("Acer rubrum", "Acer rubrum", "Betula lenta", "Betula lenta", 
         "Prunus serotina", "Prunus serotina")
spp
```

Notice that elements of a vector are separated by commas when using the `c()` function to create a vector. Also notice that character values are placed inside quotation marks.

The `c()` function also can be used to add to an existing vector. For example, if a seventh tree were included in the data set, and its DBH was 13 inches, the existing vectors could be modified as follows.

```{r}
dbh <- c(dbh, 13)
spp <- c(spp, "Acer rubrum")
dbh
spp
```

### Types, Conversion, and Coercion

Clearly it is important to distinguish between different types of vectors. For example, it makes sense to ask R to calculate the mean of the DBH stored in `dbh`, but does not make sense to ask R to compute the mean of the species stored in `spp`. Vectors in R may have one of six different "types": character, double, integer, logical, complex, and raw. Only the first four of these will be of interest below, and the distinction between double and integer will not be of great import. To illustrate logical vectors, imagine the field technician who measured the trees also indicated if the tree was acceptable growing stock (`ags`) and the call was coded as TRUE if the tree was acceptable and FALSE if the tree was not acceptable.

```{r}
typeof(dbh)
typeof(spp)
ags <-    c(TRUE, TRUE, FALSE, TRUE, FALSE, FALSE, TRUE)
ags
typeof(ags)
```

It may be surprising to see the DBH variable `dbh` is of type `double`, even though its values are all integers. By default R creates a double type vector when numeric values are given via the `c()` function.

When it makes sense, it is possible to convert vectors to a different type. Consider the following examples.

```{r}
dbh.int <- as.integer(dbh)
dbh.int
typeof(dbh.int)
dbh.char <- as.character(dbh)
dbh.char
ags.double <- as.double(ags)
ags.double
spp.oops <- as.double(spp)
spp.oops
sum(ags)
```

The integer version of `dbh` doesn't look any different, but it is stored differently, which can be important both for computational efficiency and for interfacing with other languages such as C++. As noted above, however, we will not worry about the distinction between integer and double types. Converting `dbh` to character goes as expected---the character representation of the numbers replace the numbers themselves. Converting the logical vector `ags` to double is pretty straightforward too---`FALSE` is converted to zero, and `TRUE` is converted to one. Now think about converting the character vector `spp` to a numeric double vector. It's not at all clear how to represent "Acer rubrum" as a number. In fact in this case what R does is to create a double vector, but with each element set to `NA`, which is the representation of missing data ^[Missing data will be discussed in more detail later in the chapter.]. Finally consider the code `sum(ags)`. Now `ags` is a logical vector, but when R sees that we are asking to sum this logical vector, it automatically converts it to a numerical vector and then adds the zeros and ones representing FALSE and TRUE.

R also has functions to test whether a vector is of a particular type.
```{r}
is.double(dbh)
is.character(dbh)
is.integer(dbh.int)
is.logical(ags)
```

#### Coercion

Consider the following examples.

```{r}
xx <- c(1, 2, 3, TRUE)
xx
yy <- c(1, 2, 3, "dog")
yy
zz <- c(TRUE, FALSE, "cat")
zz
dbh+ags
```

Vectors in R can only contain elements of one type. If more than one type is included in a `c()` function, R silently *coerces* the vector to be of one type. The examples illustrate the hierarchy---if any element is a character, then the whole vector is character. If some elements are numeric (either integer or double) and other elements are logical, then the whole vector is numeric. Note what happened when R was asked to add the numeric vector `dbh` to the logical vector `ags`. The logical vector was silently coerced to be numeric, so that FALSE became zero and TRUE became one, and then the two numeric vectors were added.

### Accessing Specific Elements of Vectors

To access and possibly change specific elements of vectors, refer to the position of the element in square brackets. For example, `dbh[4]` refers to the fourth element of the vector `dbh`. Note that R starts the numbering of elements at 1, i.e., the first element of a vector `x` is `x[1]`.

```{r}
dbh
dbh[5]
dbh[1:3]
length(dbh)
dbh[length(dbh)]
dbh[]
dbh[3] <- 202 ##crazy big tree
dbh
dbh[1:3] <- c(16,8,2)
dbh
```

Note that including nothing in the square brackets results in the whole vector being returned. We can also assign values to vectors by accessing the position(s) where the new values will be assigned. For example, in the above code chunk `dbh[3]` is changed to 202, then the values in the first three elements of `dbh` are changed to 10, 8, and 2, respectively.

Negative numbers in the square brackets tell R to omit the corresponding value. And a zero as a subscript returns nothing (more precisely, it returns a length zero vector of the appropriate type).

```{r, error = TRUE}
dbh[-3]
dbh[-length(dbh)]
fewer.dbh <- dbh[-c(1,3,5)]
fewer.dbh
dbh[0]
dbh[c(0,2,1)]
dbh[c(-1, 2)]
```

Note that mixing zero and other nonzero subscripts is allowed, but mixing negative and positive subscripts is not allowed.

What about the (usual) case where we don't know the positions of the elements we want? For example possibly we want the DBH of all acceptable growing stock trees in the data. Later we will learn how to subset using logical indices, which is a very powerful way to access desired elements of a vector ^[We had a prelude to this in the temperature data exercise.].

## Factors
Categorical variables such as `spp` can be represented as character vectors. In many cases this simple representation is sufficient. Consider, however, two other categorical variables, one representing crown class `S`, `I`, `C`, and `D` (i.e., **S**uppressed, **I**ntermediate, **C**odominate, **D**ominant), and another representing grade of the first log via categories `Grade 1`, `Grade 2`, and `Grade 3`. Suppose that for the small data set considered here, all trees are either dominant or codominant crown class. If we just represented the variable via a character vector, there would be no way to know that there are two other categories, representing suppressed and intermediate, because they happen to not be present in the data set. In addition, for the log grade variable the character vector representation does not explicitly indicate that there is an ordering of the levels.

Factors in R provide a more sophisticated way to represent categorical variables. Factors explicitly contain all possible levels, and allow ordering of levels.

```{r}
crown.class <- c("D", "D", "C", "D", "D", "C", "C")
grade <- c("Grade 1", "Grade 1", "Grade 3", "Grade 2", "Grade 3", "Grade 3", "Grade 2")
crown.class
grade
crown.class <- factor(crown.class, levels=c("S", "I", "C", "D"))
crown.class
grade <- factor(grade, levels=c("Grade 1", "Grade 2", "Grade 3"), ordered = TRUE)
grade
```

In the factor version of crown class the levels are explicitly listed, so it is clear that the two included levels are not all the possible levels. In the factor version of log grade, the ordering is explicit as well.

In many cases the character vector representation of a categorical variable is sufficient and easier to work with. In this book factors will not be used extensively. It is important to note that R often by default creates a factor when character data are read in, and sometimes it is necessary to use the argument `stringsAsFactors = FALSE` to explicitly tell R not to do this. This will be seen below when data frames are introduced.

## Names of objects in R

Continuing the discussion about code quality from Section \@ref(quality-of-r-code), there are a few hard and fast restrictions on the names of objects (such as vectors) in R. Note also that there are good practices, and things to avoid.

From the help page for `make.names` in R,  the name of an R object is "syntactically valid" if the name "consists of letters, numbers and the dot or underline characters and starts with a letter or the dot not followed by a number" and is not one of the "reserved words" in R such as `if`, `TRUE`, `function`, etc.  For example, `c45t.le_dog`  and `.ty56` are both syntactically valid (although not very good names) while `4DislikeCats` and `log#@sparty` are not.

A few important comments about naming objects follow:

1. It is important to be aware that names of objects in R are case-sensitive, so `dbh` and `DBH` do not refer to the same object.

```{r, error = TRUE}
dbh
DBH
```

2. It is unwise to create an object with the same name as a built-in R object such as the function `c` or the function `mean`. In earlier versions of R this could be somewhat disastrous, but even in current versions, it is definitely not a good idea!

3. As much as possible, choose names that are informative. When creating a variable you may initially remember that `x` contains DBH and `y` contains crown class, but after a few hours, days, or weeks, you probably will forget this. Better options are `dbh` and `crown.class`.
4. As much as possible, be consistent in how you name objects. In particular, decide how to separate multi-word names. Some options include:

    + Using case to separate: `CrownClass` or `crownClass` for example
    + Using underscores to separate: `crown_class` for example
    + Using a period to separate: `crown.class` for example

## Missing Data, Infinity, etc.

Most real-world data sets have variables where some observations are missing. In longitudinal studies of tree growth (i.e., where trees are measured over time), it is common that trees die or cannot be located in subsequent remeasurements. Statistical software should be able to represent missing data and to analyze data sets in which some data are missing.

In R, the value `NA` is used for a missing data value. Since missing values may occur in numeric, character, and other types of data, and since R requires that a vector contain only elements of one type, there are different types of `NA` values. Usually R determines the appropriate type of `NA` value automatically. It is worth noting the default type for `NA` is logical, and that `NA` is NOT the same as the character string `"NA"`.

```{r}
missingCharacter <- c("dog", "cat", NA, "pig", NA, "horse")
missingCharacter
is.na(missingCharacter)
missingCharacter <- c(missingCharacter, "NA")
missingCharacter
is.na(missingCharacter)
allMissing <- c(NA, NA, NA)
typeof(allMissing)
```

How should missing data be treated in computations, such as finding the mean or standard deviation of a variable? One possibility is to return `NA`. Another is to remove the missing value(s) and then perform the computation.
```{r, prompt=TRUE}
mean(c(1,2,3,NA,5))
mean(c(1,2,3,NA,5), na.rm=TRUE)
```

As this example shows, the default behavior for the `mean()` function is to return `NA`. If removal of the missing values and then computing the mean is desired, the argument `na.rm` is set to `TRUE`. Different R functions have different default behaviors, and there are other possible actions. Consulting the help for a function provides the details.

### Infinity and NaN

What happens if R code requests division by zero, or results in a number that is too large to be represented? Here are some examples.

```{r, prompt = TRUE}
x <- 0:4
x
1/x
x/x
y <- c(10, 1000, 10000)
2^y
```

`Inf` and `-Inf` represent infinity and negative infinity (and numbers which are too large in magnitude to be represented as floating point numbers). `NaN` represents the result of a calculation where the result is undefined, such as dividing zero by zero. All of these are common to a variety of programming languages, including R.

## Data Frames 

Commonly data is rectangular in form, with variables as columns and cases as rows. Continuing with the species, DBH, and acceptable growing stock data, each of those variables would be a column of the data set, and each tree's measurements would be a row. In R, such data are represented as a *data frame*.

```{r, prompt=TRUE, tidy=FALSE}
trees <- data.frame(Spp=spp, Dbh=dbh, Ags=ags, 
                    stringsAsFactors=FALSE)
trees
names(trees)
colnames(trees)
names(trees) <- c("species", "DBH", "good.stock")
colnames(trees)
rownames(trees)
names(trees) <- c("spp", "dbh", "ags")
dim(trees)
```

The `data.frame` function can be used to create a data frame (although it's more common to read a data frame into R from an external file, something that will be introduced later). The names of the variables in the data frame are given as arguments, as are the vectors of data that make up the variable's values. The argument `stringsAsFactors=FALSE` asks R not to convert character vectors into factors, which R does by default, to the dismay of many users. Names of the columns (variables) can be extracted and set via either `names` or `colnames`. In the example, the variable names are changed to `species, DBH, good.stock` and then changed back to what I like better `spp, dbh, ags` in this way. Rows can be named also. In this case since specific row names were not provided the default row names of `"1", "2"`, etc. are used. Finally, I take a look at the data frame's dimensions (where the `dim` function returns a vector comprised of number of rows and number of columns, respectively). Also, try the functions `nrow` and `ncol` on the data frame and see what happens.

In the next example a built-in R data set called `Loblolly` is made available by the `data` function, and then the first and last six rows are displayed using `head` and `tail`.

```{r, prompt=TRUE}
data("Loblolly")
head(Loblolly)
tail(Loblolly)
```

Note the `Loblolly` data frame has row names that are not ordered (which really doesn't matter) and simply suggests the data set author might have subset these data from a larger data set or sorted them by a variable, e.g., `height` or `age`. Row names can be generally ignored (unless they hold some specific meaning). Find out more about the Loblolly data set by running `?Loblolly` on the command line or, equivalently, looking it up in the RStudio's search window on the help tab. 

### Accessing specific elements of data frames

Data frames are two-dimensional, so to access a specific element (or elements) we need to specify both the row and column indices.
```{r, prompt = TRUE}
Loblolly[1,3]
Loblolly[1:3, 3]
Loblolly[1:3, 2:3]
Loblolly[,1]
```

Note that `Loblolly[,1]` returns ALL elements in the first column. This agrees with the behavior for vectors, where leaving a subscript out of the square brackets tells R to return all values. In this case we are telling R to return all rows, and the first column.

As we have seen in class, we can also access the columns (or rows) using their names.
```{r, prompt=TRUE}
Loblolly[1:4,"height"]
Loblolly[1:4,c("age","Seed")]
```

For a data frame there is another way to access specific columns, using the `$` notation.
```{r, prompt=TRUE, error = TRUE}
Loblolly$height
Loblolly$age
height
age
```

Notice that typing the variable name, such as `height`, without the name of the data frame (and a dollar sign) as a prefix, does not work. This is sensible. There may be several data frames that have variables named `height`, and just typing `height` doesn't provide enough information to know which is desired.

## Lists

The third main data structure we will work with is a list. Technically a list is a vector, but one in which elements can be of different types. For example a list may have one element that is a vector, one element that is a data frame, and another element that is a function. Consider designing a function that fits a simple linear regression model to two quantitative variables. We might want that function to compute and return several things such as

+ The fitted slope and intercept (a numeric vector with two components)
+ The residuals (a numeric vector with $n$ components, where $n$ is the number of data points)
+ Fitted values for the data (a numeric vector with $n$ components, where $n$ is the number of data points)
+ The names of the dependent and independent variables (a character vector with two components)

In fact R has a function, `lm`, which does this (and much more).

```{r, prompt=TRUE}
htAgeLinMod <- lm(height ~ age, data=Loblolly)
mode(htAgeLinMod)
names(htAgeLinMod)
htAgeLinMod$coefficients
tail(htAgeLinMod$residuals)
```

The `lm` function returns a list (which in the code above has been assigned to the object `htAgeLinMod`)^[The `mode` function returns the the type or storage mode of an object.]. One component of the list is the length `r length(htAgeLinMod$coefficients)` vector of coefficients, while another component is the length `r length(htAgeLinMod$residuals)` vector of residuals. The code also illustrates that named components of a list can be accessed using the dollar sign notation, as with data frames.

The `list` function is used to create lists.
```{r prompt=TRUE}
temporaryList <- list(first=dbh, second=trees, pickle=list(a = 1:10, b=trees))
temporaryList
```

Here, for illustration, I assembled a list to hold some of the R data structures we have been working with in this chapter. The first list element, named `first`, holds the `dbh` vector we created in Section \@ref(vectors). The second list element, named `second`, holds the `trees` data frame. The third list element, named `pickle`, holds a list with elements named `a` and `b` that hold a vector of values 1 through 10, and another copy of the `trees` data set, respectively. As this example shows, a list can contain another list.

### Accessing specific elements of lists

We already have seen the dollar sign notation works for lists. In addition, the square bracket subsetting notation can be used. But with lists there is an added somewhat subtle wrinkle---using either single or double square brackets.

```{r, prompt=TRUE}
temporaryList$first
mode(temporaryList$first)
temporaryList[[1]]
mode(temporaryList[[1]])
temporaryList[1]
mode(temporaryList[1])
```

Note the dollar sign and double bracket notation return a numeric vector, while the single bracket notation returns a list. Notice also the difference in results below.

```{r, prompt=TRUE}
temporaryList[c(1,2)]
temporaryList[[c(1,2)]]
```

The single bracket form returns the first and second elements of the list, while the double bracket form returns the second element in the first element of the list. Generally, do not put a vector of indices or names in a double bracket, you will likely get unexpected results. See, for example, the results below ^[Try this example using only single brackets$\ldots$ it will return a list holding elements `first`, `second`, and `pickle`.]. 

```{r, prompt=TRUE, error = TRUE}
temporaryList[[c(1,2,3)]]
```

So, in summary, there are two main differences between using the single bracket `[]` and double bracket `[[]]`. First, the single bracket will return a list that holds the object(s) held at the given indices or names placed in the bracket, whereas the double brackets will return the actual object held at the index or name placed in the innermost bracket. Put differently, a single bracket can be used to access a range of list elements and will return a list, while a double bracket can only access a single element in the list and will return the object held at the index.

## Subsetting with Logical Vectors

Consider the `Loblolly` data frame. How can we access only those trees with heights more than 50 m? How can we access the age of those trees taller than 50 m? How can we compute the mean height of all trees from seed source 301? The data set is small enough that it would not be too onerous to extract the values by hand. But for larger or more complex data sets, this would be very difficult or impossible to do in a reasonable amount of time, and would likely result in errors.

R has a powerful method for solving these sorts of problems using a variant of the subsetting methods that we already have learned. When given a logical vector in square brackets, R will return the values corresponding to `TRUE`. To begin, focus on the `dbh` and `spp` vectors created in Section \@ref(vectors).

The R code `dbh > 15` returns `TRUE` for each value of `dbh` that is more than 15, and a `FALSE` for each value of `dbh` that is less than or equal to 15. Similarly `spp == "Betula lenta"` returns `TRUE` or `FALSE` depending on whether an element of `spp` is equal to `Betula lenta`.

```{r, prompt = TRUE}
dbh
dbh > 15
spp[dbh > 15]
dbh[dbh > 15]
spp == "Betula lenta"
dbh[spp == "Betula lenta"]
```

Consider the lines of R code one by one.

+ `dbh` instructs R to display the values in the vector `dbh`.
+ `dbh > 15` instructs R to check whether each value in `dbh` is greater than 15, and to return `TRUE` if so, and `FALSE` otherwise.
+ The next line, `spp[dbh > 15]`, does two things. First, inside the square brackets, it does the same thing as the second line: it returns `TRUE` or `FALSE` depending on whether a value of `dbh` is or is not greater than 15. Second, each element of `spp` is matched with the corresponding `TRUE` or `FALSE` value, and is returned if and only if the corresponding value is `TRUE`. For example the first value of `spp` is `r spp[1]`. Since the first `TRUE` or `FALSE` value is `TRUE`, the first value of `spp` is returned. So this line of code returns the species names for all trees with DBH greater than 15; hence, the first and the fourth values of `spp` are returned. 
+ The fourth line of code, `dbh[dbh > 15]`, again begins by returning `TRUE` or `FALSE` depending on whether elements of `dbh` are larger than 15. Then those elements of `dbh` corresponding to `TRUE` values are returned. So this line of code returns the DBH of all trees whose DBH is greater than 15.
+ The fifth line returns `TRUE` or `FALSE` depending on whether elements of `spp` are equal to `Betula lenta` or not.
+ The sixth line returns the DBH of those all Betula lenta trees.

There are six comparison operators in R, `>, <, >=, <=, ==, !=`. Note that to test for equality a "double equals sign"" is used, while `!=` tests for inequality.


