# Working with Data Sources

Bringing data into R, exporting data from R in a form that is readable by other software, cleaning and reshaping data, and other data manipulation tasks are an important and often overlooked component of data science. The book @SpectorDataManipulation, while a few years old, is still an excellent reference for data-related issues. And the *R Data Import/Export* manual, available online at https://cran.r-project.org/doc/manuals/r-release/R-data.html, is an up-to-date (and free) reference on importing a wide variety of datasets into R and on exporting data in various forms.

## Reading Data into R

Data come in a dizzying variety of forms. It might be in a proprietary format such as a .xlsx Excel file, a .sav SPSS file, or a .mtw Minitab file. It might be structured using a [relational model](https://en.wikipedia.org/wiki/Relational_model), for example, the USDA Forest Service [Forest Inventory and Analysis database](https://www.fia.fs.fed.us/tools-data/). It might be a data-interchange format such as [JSON](http://www.json.org/) (JavaScript Object Notation),  or a markup language format such as [XML](https://en.wikipedia.org/wiki/XML) (Extensible Markup Language), perhaps with specialized standards for describing ecological information, see, e.g., [EML](https://en.wikipedia.org/wiki/Ecological_Metadata_Language) (Ecological Metadata Language). Both XML and EML are common data metadata formats (i.e., data that provides information about other data). Fortunately many datasets are (or can be) saved as plain text files, and most software can both read and write such files, so our initial focus will be on reading plain text files into R and saving data from R in plain text format. RStudio provides a handy [data import cheat sheet](http://www.rstudio.com/resources/cheatsheets) for many of the read functions detailed in this section.

The `foreign` R package provides functions to directly read data saved in some of the proprietary formats into R, which is sometimes unavoidable, but if possible it is good to save data from another package as plain text and then read this plain text file into R. 

The function `read.table()` and its offshoots such as `read.csv()` are used to read in rectangular data from a text file. For example, the file `FEF-trees.csv` described in Section \@ref(fef) contains biomass information for trees measured on plots within different watersheds across years. Below are the first seven rows and columns of that file:
```
watershed,year,plot,species,dbh_in,height_ft,stem_green_kg
3,1991,29,Acer rubrum,6,48,92.2
3,1991,33,Acer rubrum,6.9,48,102.3
3,1991,35,Acer rubrum,6.4,48,124.4
3,1991,39,Acer rubrum,6.5,49,91.7
3,1991,44,Acer rubrum,7.2,51,186.2
3,1992,26,Acer rubrum,3.1,40,20.8
```

As is evident, the first line of the file contains the names of the variables, separated (delimited) by commas. Each subsequent line contains information about each tree's location (watershed code), measurement year, plot (within watershed), species, DBH, height, and weight of various components of the tree (e.g., `stem_green_kg` is the green weight of the stem in kilograms).^[The metadata for the FEF are at http://www.fs.usda.gov/rds/archive/Product/RDS-2016-0016. For convenience I renamed the data file from the original `felled_tree_biomass.csv` to `FEF-trees.csv`.]

This file is accessible at http://blue.for.msu.edu/FOR472/data/FEF-trees.csv. The `read.table()` function is used to read these data into an R data frame.

```{r}
fef.file <- "http://blue.for.msu.edu/FOR472/data/FEF-trees.csv"
fef.trees <- read.table(file = fef.file, header = TRUE, sep = ",")
head(fef.trees[,1:6])
```

The arguments used in this call to `read.table()` include:

+ `file = fef.file` tells R the location of the file. In this case the string `http://blue.for.msu.edu/FOR472/data/FEF-trees.csv` giving the location is rather long, so it was first assigned to the object `fef.file`.
+ `header = TRUE` tells R the first line of the file gives the names of the variables.
+ `sep = ","` tells R that a comma separates the fields in the file.

The function `read.csv()` is the same as `read.table()` except the default separator is a comma, whereas the default separator for `read.table()` is whitespace. 

The file `FEF-trees.tsv` contains the same data, except a tab is used in place of a comma to separate fields. The only change needed to read in the data in this file is in the `sep` argument (and of course the `file` argument, since the data are stored in a different file):

```{r}
fef.file <- "http://blue.for.msu.edu/FOR472/data/FEF-trees.tsv"
fef.trees <- read.table(file = fef.file, header = TRUE, sep = "\t")
head(fef.trees[,1:6])
```

File extensions, e.g., `.csv` or `.tsv`, are naming conventions only and are there to remind us how the columns are delimited, i.e., they have no influence on R's file read functions.

A third file, [`FEF-trees.txt`](http://blue.for.msu.edu/FOR472/data/FEF-trees.txt), contains the same data, but also contains a few lines of explanatory text above the names of the variables. It also uses whitespace rather than a comma or a tab as a delimiter. Here are the first several lines of the file and six columns of data.

```
This file includes felled tree biomass by tree component for hardwood 
species sampled on the Fernow Experimental Forest (FEF), West Virginia. 
A total of 88 trees were sampled from plots within two watersheds. 

"watershed" "year" "plot" "species" "dbh_in" "height_ft" "stem_green_kg"
3 1991 29 "Acer rubrum" 6 48 92.2
3 1991 33 "Acer rubrum" 6.9 48 102.3
3 1991 35 "Acer rubrum" 6.4 48 124.4
3 1991 39 "Acer rubrum" 6.5 49 91.7
3 1991 44 "Acer rubrum" 7.2 51 186.2
3 1992 26 "Acer rubrum" 3.1 40 20.8
```

Notice that in this file column (variable) names are put inside of quotation marks. Also, variable values that are characters are also quoted. This is necessary because character strings could include whitespace, and hence R would assume these are column delimiters. 

To read in this file we need to tell R to skip the first four lines above the header and also that whitespace is the delimiter. The `skip` argument handles the first, and the `sep` argument the second.

```{r}
fef.file <- "http://blue.for.msu.edu/FOR472/data/FEF-trees.txt"
fef.trees <- read.table(file = fef.file, header = TRUE, sep = " ", skip=4)
head(fef.trees[,1:6])
```

For fun, see what happens when the `skip` argument is left out of the `read.table` call.

## Reading Data with missing observations

Missing data are represented in many ways. Sometimes missing data are just that, i.e., the place where they should be in the file is blank. Other times specific numbers such as $-9999$ or specific symbols are used. The `read.table()` function has an argument `na.string` which allows the user to specify how missing data are indicated in the source file.

The site http://www.wunderground.com/history/ makes weather data available for locations around the world, and for dates going back to 1945. The file `WeatherKLAN2014.csv` contains weather data for Lansing, Michigan for the year 2014. Here are the first few lines of that file:

```
EST,Max TemperatureF,Min TemperatureF, Events
1/1/14,14,9,Snow
1/2/14,13,-3,Snow
1/3/14,13,-11,Snow
1/4/14,31,13,Snow
1/5/14,29,16,Fog-Snow
1/6/14,16,-12,Fog-Snow
1/7/14,2,-13,Snow
1/8/14,17,-1,Snow
1/9/14,21,2,Snow
1/10/14,39,21,Fog-Rain-Snow
1/11/14,41,32,Fog-Rain
1/12/14,39,31,
```

Look at the last line, and notice that instead of an event such as `Snow` or `Fog-Snow` there is nothing after the comma. This observation is missing, but rather than using an explicit code such as `NA`, the site just leaves that entry blank. To read these data into R we will supply the argument `na.string = ""` which tells R the file indicates missing data by leaving the appropriate entry blank.

```{r, tidy = FALSE}
u.weather <- "http://blue.for.msu.edu/FOR875/data/WeatherKLAN2014.csv"
WeatherKLAN2014 <- read.csv(u.weather, header=TRUE, 
                            stringsAsFactors = FALSE, na.string = "")
WeatherKLAN2014[1:15,]
```

Also, in the code above, notice I use `stringsAsFactors = FALSE` to prevent the character variables to be converted to factors, which R does by default (which can be quite annoying). 
